---
title: "36-469 Homework 03, Fall 2021"
author: "Leandro Lopez Leon (lslopez)"
date: "Monday, Oct. 11, 2021" 
output:
  pdf_document:
    toc: no
  html_document:
    toc: true
    toc_float: true
    theme: spacelab
---

```{r wrap-hook,echo=FALSE}
    library(knitr)
    library(tidyverse)
    hook_output = knit_hooks$get('output')
    knitr::opts_chunk$set(echo = F, warning = F, message = F)
    knit_hooks$set(output = function(x, options) {
      # this hook is used only when the linewidth option is not NULL
      if (!is.null(n <- options$linewidth)) {
        x = knitr:::split_lines(x)
        # any lines wider than n should be wrapped
        if (any(nchar(x) > n)) x = strwrap(x, width = n)
        x = paste(x, collapse = '\n')
      }
      hook_output(x, options)
    })
```

## Q1
**A**.  

```{r}
    library(tidyverse)
    library(gridExtra)
    snp_data_original <- read.csv("https://raw.githubusercontent.com/xuranw/469_public/master/data/snp_data.csv", row.names = 1)
    heart_disease <- snp_data_original[,1]
    snp_data <- snp_data_original[,-1]
    logistic_regression  <- function(covariate, dataset) {
        glm_output <- stats::glm(as.formula(paste("heart_disease ~", covariate)), data = dataset, family = stats::binomial)
        summary(glm_output)$coefficients[2,4]
    }
    p_values <- lapply(names(snp_data), logistic_regression, dataset = snp_data_original)
    plot_diagnostics <- function(pvalues, correction = T) {
        if (correction) {
            title_left <- "Histogram of p-values without correction for ancestry"
            title_right <- "QQ plot of p-values without correction for ancestry"
        } else {
            title_left <- "Histogram of p-values with correction for ancestry"
            title_right <- "QQ plot of p-values with correction for ancestry"
        }
        p_values_df <- data.frame(unlist(p_values))
        p_left  <-  ggplot(p_values_df, aes(x = unlist.p_values.)) + geom_histogram(bins = 20)  +
                            geom_vline(xintercept = 0.05, color = "red", linetype = "dotted") +
                            labs(title = title_left) +
                            xlab("P-value bin") + ylab("Frequency")

        n.data <- length(p_values)
        scaled_p_values <- sort(-1 * log10(unlist(p_values)))
        theoretical <- sort(-1 * log10(runif(n.data)))
        scaled_p_values_df <- data.frame(cbind(scaled_p_values, theoretical))
        p_right <- ggplot(scaled_p_values_df) + geom_point(aes(x = theoretical, y = scaled_p_values)) +
                    geom_abline(intercept = 0, slope = 1, color = "red") + 
                    labs(title = title_right) +
                    xlab("Expected distribution (-Log10 of p-value)") + 
                    ylab("Observed distribution (-Log10 of p-value)")
        list(p_left, p_right)
    }
    L <- plot_diagnostics(p_values)
    grid.arrange(L[[1]], L[[2]], ncol = 2)
```

Observe that the distribution of p-values for our individual regressions suggests that the overwhelming 
majority of the SNP's are significant predictors to our target (`heart_disease`). However, this diverges from what we expect, 
namely that most SNP's are unrelated to `heart_disease`, which is expressed mathematically as a uniform prior over the space of coefficients' 
p-values (see QQ-plot on the right). This discrepancy puts into question this naive approach. We must somehow account for the confounders in our
data.

**B**.  

```{r}
    pca_res <- stats::prcomp(snp_data, center = T, scale. = T)
    normalized_evs <- pca_res$sdev^2 / sum(pca_res$sdev^2)
    evs_df <- data.frame(cbind(100 * normalized_evs, seq(from = 1, to = length(normalized_evs), by = 1)))
    ggplot(evs_df, aes(x = X2, y = X1)) + geom_point() + labs(title = "Scree plot") +
        xlab("Components") + ylab("% of variance accounted for")
```

**C**.  

```{r}
    pc1and2 <- pca_res$rotation[,c(1,2)]
    coordinates <- data.frame(cbind(as.matrix(snp_data) %*% pc1and2, heart_disease))
    ggplot(coordinates, aes(x = PC1, y = PC2, color = as.factor(heart_disease))) +
        geom_point() +
        labs(x = "Principal Component 1",
             y = "Principal Component 2",
             color = "Has CHD?", 
             title = "PCA results")
```
**D**.  

```{r}
    logistic_regression_withPC  <- function(covariate, dataset) {
        glm_output <- stats::glm(as.formula(paste("heart_disease ~ PC1 + PC2 +", covariate)), data = dataset, family = stats::binomial)
        summary(glm_output)$coefficients[4,4]
    }
    snp_data_PC <- snp_data_original %>% mutate(PC1 = coordinates$PC1, PC2 = coordinates$PC2)
    p_values <- lapply(names(snp_data), logistic_regression_withPC, dataset = snp_data_PC)
    L <- plot_diagnostics(p_values, correction = F)
    grid.arrange(L[[1]], L[[2]], ncol = 2)
```
We note that the empirical distribution of the p-values looks significantly more balanced than what we saw in Q1 and that 
it tracks the expected distribution (uniform over the $[0,1]$ interval) much more closely as evidence in the QQ plot. Including ancestry in our model (via $PC1$ and $PC2$) 
has weakened the statistical significance between each individual SNP and our response (CHD). In effect, we have taken into account this confounder in our analysis. 

## Q2
```{r}
    source("https://raw.githubusercontent.com/xuranw/469_public/master/hw3/hw3_functions.R")
    set.seed(10); dat <- generate_data()
    first_pc <- stats::prcomp(dat)$x[,1]
```

**A**.  
```{r}
    S <- cor(dat)
    eigen_res <- eigen(S)
    first_pc_eigen <- -1 * dat %*% eigen_res$vectors[,1]
    svd_res <- svd(dat)
    first_pc_svd <- (dat %*% svd_res$v[,1])
```

Below we show that the principal components computed above are equal 

```{r, echo = T}
    all.equal(as.vector(first_pc_eigen), first_pc, as.vector(first_pc_svd))
```

And we report the value of $\lambda^{(1)}$

```{r}
    eigen_res$values[1]
```

**B**.  

```{r}
    compute_variance <- function(X, w) {
        var(apply(X, 1, function(xi) xi %*% w))
    }

    compute_reconstruction <- function(X, w) {
        wwX <- as.data.frame(t(X %*% (w %*% t(w))))
        X <- as.data.frame(t(X))
        sum(mapply(function(xi, wwxi) {sum((xi - wwxi)^2)}, X, wwX))^(1 / 2)
    }

    plot_fig5_and6 <- function(coordinates, variance, reconstruction, PCA = T) {
        if (PCA) { 
            title_pca <- paste(paste("PCA: Variance", round(variance, digits = 2), ",\n"),
                               paste("Reconstruction",  round(reconstruction, digits = 2)))
            color_fig_dot <- "red"
            label_fig_dot <- "(PCA) Projection"
        } else { 
            title_pca <- paste(paste("Random Projection: Variance", round(variance, digits = 2), ",\n"),
                               paste("Reconstruction",  round(reconstruction, digits = 2)))
            color_fig_dot <- "green"
            label_fig_dot <- "(Random) Projection"
        }

        ggplot(coordinates, aes(color = as.factor(X3))) +
            geom_point(aes(x = X1, y = X2)) +
            labs(x = "Dimension 1",
                 y = "Dimension 2",
                 color = paste("Original = 0, \n", paste(label_fig_dot, "= 1")),
                 title = title_pca) +
            scale_color_manual(values = c("black", color_fig_dot))
    }

    pc1and2 <- eigen_res$vectors[,c(2,1)]
    red_dots  <- cbind(dat %*% (-1*eigen_res$vectors[,1] %*% t(-1*eigen_res$vectors[,1])), rep(1, length(dat[,1])))
    black_dots <- cbind(dat %*% pc1and2, rep(0, length(dat[,1])))
    coordinates <- data.frame(rbind(black_dots, red_dots))
    variance <- compute_variance(dat, -1*eigen_res$vectors[,1])
    reconstruction <- compute_reconstruction(dat, -1*eigen_res$vectors[,1])
    plot_fig5_and6(coordinates, variance, reconstruction)
```

**C**.  

```{r}
    w_random <- generate_random_direction()
    green_dots  <- cbind(dat %*% (w_random %*% t(w_random)), rep(1, length(dat[,1])))
    coordinates <- data.frame(rbind(black_dots, green_dots))
    variance <- compute_variance(dat, w_random)
    reconstruction <- compute_reconstruction(dat, w_random)
    plot_fig5_and6(coordinates, variance, reconstruction, PCA = F)
```
