---
title: "36-469 Homework 06, Fall 2021"
author: "Leandro Lopez Leon (lslopez)"
date: "Monday, Nov. 22, 2021 (1 Day Late)" 
output:
  pdf_document:
    toc: no
  html_document:
    toc: true
    toc_float: true
    theme: spacelab
---

```{r wrap-hook,echo=FALSE}
    library(knitr)
    library(tidyverse)
    hook_output = knit_hooks$get('output')
    knitr::opts_chunk$set(echo = F, warning = F, message = F)
    knit_hooks$set(output = function(x, options) {
      # this hook is used only when the linewidth option is not NULL
      if (!is.null(n <- options$linewidth)) {
        x = knitr:::split_lines(x)
        # any lines wider than n should be wrapped
        if (any(nchar(x) > n)) x = strwrap(x, width = n)
        x = paste(x, collapse = '\n')
      }
      hook_output(x, options)
    })
```

## Q0 (TODO)
I'm interested in using spectar partioning to cluster SNN graphs. This type of analysis has been done in the following paper: C. Gkantsidis, M. Mihail, and E. Zegura, _Spectral analysis of Internet
topologies_ in Proceedings of the 22nd Annual INFOCOM Conference, 2003, pp. 364–374. I will be working alone

## Q1
```{r}
    library(tidyverse)
    library(gridExtra)
    library(PMA)
```

```{r}
    tmp <- read.csv("../469_public/hw6/darmanis_preprocessed.csv")
    expr_mat <- as.matrix(tmp[,-1])
    cell_types <- as.factor(tmp[,1])
    source("../469_public/hw6/hw6_functions.R")
    dim(expr_mat); length(cell_types)
    table(cell_types)
    expr_mat[1:5, 1:5]
```
**A**.  

```{r}
    expr_numeric <- expr_mat[, colnames(expr_mat) != 'cell_type']
    m <- apply(expr_numeric, c(1,2), function(cell) as.numeric(trimws(cell)))
    m_df <- as_tibble(m)
```

```{r}
    library_norm <- function(cell) {
        cell_total <- sum(cell)
        norm_ <- function(y_i) {
                    tryCatch(
                        log2((1e4 * (y_i / cell_total)) + 1),
                        error = function(e) {})
                }
        res <- as_tibble(purrr::map(cell, norm_))
        names(res) <- names(cell)
        return(res)
    }

    norm_l <- vector("list", length(nrow(m_df)))
    for (i in 1:nrow(m_df)) {
        norm_l[[i]] <- library_norm(m_df[i, ])
    }
    norm_df <- dplyr::bind_rows(norm_l) %>% map_df(., scale) %>% mutate(cell_types = cell_types) %>%
               relocate(cell_types)
    norm_df[1:5, 1:6]
```
\pagebreak

**B**.  

```{r}
    norm_df_numeric <- norm_df %>% select(-cell_types)
    pca_res <- stats::prcomp(norm_df_numeric, center = T, scale. = T)
    tmp <- as.matrix(norm_df_numeric) %*% pca_res$rotation[, 1:4]
    expr_pca <- apply(tmp, 2, scale)
    scree_df <- tibble(x = 1:length(pca_res$sdev), y = pca_res$sdev)
    p1 <- ggplot(scree_df, aes(x = x, y = y)) + geom_point() + labs(title = "Scree plot of full data") +
        xlab("Index of Principal Components") + ylab("Square root of eigenvalues")
    coordinates <- tibble(PC1 = expr_pca[,1], PC2 = expr_pca[,2], cell_type = expr_mat[,1])
    p2 <- ggplot(coordinates, aes(x = PC1, y = PC2, color = cell_type)) +
        geom_point() +
        labs(x = "Principal Component 1",
             y = "Principal Component 2",
             title = "Visualizing data\n(True Clusters, full data)")
    grid.arrange(p1, p2, nrow = 2)
```

We see a dramatic drop in the square root of the eigenvalues after the fourth index, 
suggesting that we can capture the overwhelming majority of the variance with the 
first four principal components. Additionally, our first two principal 
components (somewhat) allow us to draw decision boundaries between the different cell types.


**C**.  

Confusion Matrix 

```{r}
    kmeans_res <- stats::kmeans(expr_numeric, centers = 4)
    cell_type <- as.factor(expr_mat[,1])
    table(kmeans_res$cluster, cell_type)
```

Misclustering rate

```{r}
    compute_misclustering_rate(kmeans_res$cluster, cell_type)
```

Note that naive K-means misclassifies many `astrocytes` and `fetal_quiescent` cell types 
by placing them into the `neurons` category. 

**D**.  

Confusion Matrix 

```{r}
    kmeans_res_pca <- stats::kmeans(expr_pca, centers = 4)
    table(kmeans_res_pca$cluster, cell_type)
```

Misclustering rate

```{r}
    compute_misclustering_rate(kmeans_res_pca$cluster, cell_type)
```

```{r}
    coordinates <- tibble(PC1 = expr_pca[, 1], PC2 = expr_pca[, 2],
                          cell_type = kmeans_res_pca$cluster)
    ggplot(coordinates, aes(x = PC1, y = PC2, color = cell_type)) +
        geom_point() +
        guides(color = "none") +
        labs(x = "Principal Component 1",
             y = "Principal Component 2",
             title = "Visualizing data\n(Estimated Clusters, full data)")
```

We note that the estimated clusters gets confused around the cloud near $(0,0)$

**E**.  

```{r, include = F}
    expr_numeric <- as.matrix(norm_df_numeric)
    spca_cv_res <- PMA::SPC.cv(expr_numeric, sumabsvs =
                               seq(1.2, sqrt(ncol(expr_numeric)) / 2,
                                   length.out = 10))

    spca_res <- PMA::SPC(expr_numeric, sumabsv = spca_cv_res$bestsumabsv1se, K = 4)
    gene_idx <- which(apply(spca_res$v, 1, function(r) !all(r == 0)))
    expr_mat_screened <- as_tibble(expr_numeric[, gene_idx]) %>% mutate(cell_types = cell_types) %>%
                         relocate(cell_types)
```

```{r}
expr_mat_screened[1:6, ]
```
\pagebreak

**F**.  

Confusion Matrix 

```{r}
    expr_mat_screened_numeric <-  expr_mat_screened %>% select(-cell_types)
    pca_res_s <- stats::prcomp(expr_mat_screened_numeric, center = T, scale. = T)
    tmp <- as.matrix(expr_mat_screened_numeric) %*% pca_res_s$rotation[, 1:4]
    expr_spca <- apply(tmp, 2, scale)
    kmeans_res_spca <- stats::kmeans(expr_spca, centers = 4)
    table(kmeans_res_spca$cluster, cell_type)
```

Misclassification rate

```{r}
    compute_misclustering_rate(kmeans_res_spca$cluster, cell_type)
```

```{r}
    first_df <- tibble(PC1 = expr_spca[,1], PC2 = expr_spca[,2], cell_type = expr_mat[,1])

    p1 <- ggplot(first_df, aes(x = PC1, y = PC2, color = cell_type)) +
        geom_point() +
        labs(x = "Principal Component 1",
             y = "Principal Component 2",
             title = "Visualizing data\n(True Clusters, full data)")

    second_df <- tibble(PC1 = expr_spca[,1], PC2 = expr_spca[,2], cell_type = kmeans_res_spca$cluster)
    p2 <- ggplot(second_df, aes(x = PC1, y = PC2, color = cell_type)) +
        geom_point() +
        guides(color = "none") +
        labs(x = "Principal Component 1",
             y = "Principal Component 2",
             title = "Visualizing data\n(Est. Clusters, full data)")
    grid.arrange(p1, p2, nrow = 2)
```

With the untransformed data, PCA's factors can give disproportionate importance to noise in the data. 
By filtering out unimportant genes before performing PCA, we guarantee that the computed factor loading are 
more meaningful, resulting in a better clustering performance since the euclidean distance is not swayed by 
irrelevant features.
