---
title: "36-469 Homework 04, Fall 2021"
author: "Leandro Lopez Leon (lslopez)"
date: "Monday, Oct. 25, 2021 (one day late)" 
output:
  pdf_document:
    toc: no
  html_document:
    toc: true
    toc_float: true
    theme: spacelab
---

```{r wrap-hook,echo=FALSE}
    library(knitr)
    library(tidyverse)
    hook_output = knit_hooks$get('output')
    knitr::opts_chunk$set(echo = F, warning = F, message = F)
    knit_hooks$set(output = function(x, options) {
      # this hook is used only when the linewidth option is not NULL
      if (!is.null(n <- options$linewidth)) {
        x = knitr:::split_lines(x)
        # any lines wider than n should be wrapped
        if (any(nchar(x) > n)) x = strwrap(x, width = n)
        x = paste(x, collapse = '\n')
      }
      hook_output(x, options)
    })
```

## Q1
```{r}
    library(tidyverse)
    library(gridExtra)
    library(hexbin)
    library(DESeq2)
    library(vsn)
    library(SummarizedExperiment)
    library(genefilter)
    library(SummarizedExperiment); library(genefilter)
    load(url("https://github.com/xuranw/469_public/raw/main/hw5/airway.RData"))
```

**A**.  

```{r}
    class(dds)
    SummarizedExperiment::colData(dds)
    count_dds  <- SummarizedExperiment::assay(dds)
    class(count_dds)
    dim(count_dds)
    colnames(count_dds)
    head(rownames(count_dds))
    rownames(count_dds)[245]
```
**1** The $(i,j)$th element of `count_dds` represents the number of 
sequence fragments which have been assigned to gene $i$ in experiment $j$ 

**2** In the output of `colData(dds)`, each subject seems to have 
their own row. For each subject we see whether they are part of 
the control or the treatment group, donor ID, cell ID 

**3** Couldn't find it `:(`

**B**.  

```{r}
    dds <- DESeq2::DESeq(dds)
    res <- DESeq2::results(dds)
    res
```
`DESeq` first estimates the normalization constant $s_j$ for all samples $j$ . Then
it estimates the count dispersion $\alpha_i$ for each gene. Then it 
finds suitable coefficients $B_i$ for a GLM of the Negative Binomial family with $log$ link function (technically estimating a posterior). Finally it tests 
the significance of these coefficients using a Walt Test. This last step is the 
most important since steps 1-3 may not have yielded parameters with 
any explanatory power. 

**C**.  

```{r}
    DESeq2::plotMA(res, ylim = c(-5, 5), main = "MA plot")
```
Each point represents a gene; blue ones represent significant ones and grey 
nonsignificant ones (Significance computed depends on coefficient value). $y$ axis
represents how much the gene's expression seems to have changed due to treatment,
while the $x$ axis represents the average normalized counts seen in training 

**D**.  

```{r}
    vsd <- DESeq2::vst(dds, blind = FALSE)
    count_vsd <- SummarizedExperiment::assay(vsd)
    vsn::meanSdPlot(count_dds)
    vsn::meanSdPlot(count_vsd)
```

Quantiles of `count_dds`
```{r}
    quantile(count_dds)
```
Quantiles of `count_vsd`
```{r}
    quantile(count_vsd)
```
**1** Note that the first distribution is heavy tailed, which means a central assumption 
of linear models (homoscedasticity) is suspect in the untransformed data. 

**2** Each hexagon in the plot displays a gene, which is color coded by its 
observed count. The $y$ axis is displaying the standard deviation of the count 
data while the $x$ axis is displaying the mean. We see a clear functional 
relationship in the first plot, whereas the second plot looks more uniform. 

**3** The right plot looks more uniform because our vst integrates over a scaled
transformation of the data where the scale is determined by the variance-mean
dependence $w(q)$

**E**

```{r}
    topVarGenes <- head(order(genefilter::rowVars(count_vsd),
                    decreasing = TRUE), 40)
    dat <- count_vsd[topVarGenes,]
    dat <- dat - rowMeans(dat)
    anno <- as.data.frame(SummarizedExperiment::colData(vsd)[,c("cell","dex")])
    pheatmap::pheatmap(dat, annotation_col = anno)
```

**1** We sort genes by count variance across the different samples and we 
pick the top 40 by this measure 

**2** The rows correspond to genes, while the columns correspond to subjects. 
Each subject is either treated of untreated and there are also different types 
of cells for each subject. The clusters can be read from root to leaf as follows:
For some level of the tree fix a node $n$, the subtrees rooted at $n$ are 
closer to each other than those subtrees which are rooted at any of $n$'s siblings 
(those nodes which are at the same level as $n$ and share the same parent)

**3**
We observe a difference in variance between untreated and treated subjects with 
the same cell attribute, which might suggest some effect related to the drug in
question. The column hierarchical clustering seems to capture this trend since 
it perfectly splits treated from non-treated subjects. The row tree has less clear
interpretation, but seems to separate genes with extreme discrepancies in treated
vs untreated, when compared with genes with more uniform variance counts 

## Q2

**A**
```{r}
   dat <- read.csv("https://raw.githubusercontent.com/xuranw/469_public/master/hw5/synthetic.csv")
    plot(dat$x, dat$y, pch = 16, xlab = "x", ylab = "y")
```
```{r}
    fit_lm <- stats::lm(y ~., data = dat)
    pred_lm <- stats::predict(fit_lm, newdata = dat, type = "response")
    est_sd <- summary(fit_lm)$sigma
    plot(dat$x, dat$y, pch = 16, ylim = range(c(pred_lm, dat$y)),
    main = "Gaussian fit", xlab = "x", ylab = "y")
    points(dat$x, pred_lm, col = "red", pch = 16, xlab = "x", ylab = "y")
    for(i in 1:length(dat$x)){
        lines(rep(dat$x[i], 2), c(pred_lm[i]-2*est_sd, pred_lm[i] + 2 * est_sd),
                                  col ="red")
    }
```
```{r}
    res <- as.vector(residuals(fit_lm))
    data.frame(ft = pred_lm, res = res) %>%
        ggplot(data = ., aes(x = ft, y = res)) +
        geom_point() + geom_smooth() + labs(ylab = "Fit", xlab = "residuals")
```
We display the residuals vs fit plot above, suggesting a functional relationship 
between the residuals and the fitted valus. This suggests that our linear model
is not capturing the complexity of our training set (if it were the residuals 
would look like white noise)

**B**
```{r}
    poisson_fit  <- glm(y ~ ., family = poisson, dat)
    pred_lm <- predict(poisson_fit, newdata = dat, type = "response")
    est_sd <- sqrt(mean(pred_lm))
    plot(dat$x, dat$y, pch = 16, ylim = range(c(pred_lm, dat$y)),
    main = "Poisson fit", xlab = "x", ylab = "y")
    points(dat$x, pred_lm, col = "red", pch = 16, xlab = "x", ylab = "y")
    for(i in 1:length(dat$x)){
        lines(rep(dat$x[i], 2), c(pred_lm[i]-2*est_sd, pred_lm[i] + 2 * est_sd),
                                  col ="red")
    }
```
```{r}
    ggplot(data = data.frame(fit = pred_lm, variance_approx = (dat$y - pred_lm)^2),
           aes(x = fit, y = variance_approx)) + geom_point() + geom_smooth()
```

Note that in the first plot, our fitted points seem to track the convexity 
of the data-generating process. However we see in the above plot of variance vs 
mean, that the relationship between the $x$ and $y$ is not linear as we expected from our model assumptions.

**C**
```{r}
    library(MASS)
    nb_fit <- glm.nb(y ~ ., data = dat)
    pred_lm <- predict(nb_fit, newdata = dat, type = "response")
    plot(dat$x, dat$y, pch = 16, ylim = range(c(pred_lm, dat$y)),
    main = "Poisson fit", xlab = "x", ylab = "y")
    points(dat$x, pred_lm, col = "red", pch = 16, xlab = "x", ylab = "y")
    for(i in 1:length(dat$x)){
        lines(rep(dat$x[i], 2), c(pred_lm[i]-0.25, pred_lm[i] + 0.25),
                                  col ="red")
    }
```
