---
title: "36-469 Homework 02, Fall 2021"
author: "Leandro Lopez Leon (lslopez)"
date: "Sunday, Sept. 26, 2021" 
output:
  pdf_document:
    toc: no
  html_document:
    toc: true
    toc_float: true
    theme: spacelab
---

```{r wrap-hook,echo=FALSE}
    library(knitr)
    library(tidyverse)
    hook_output = knit_hooks$get('output')
    knitr::opts_chunk$set(echo = F, warning = F, message = F)
    knit_hooks$set(output = function(x, options) {
      # this hook is used only when the linewidth option is not NULL
      if (!is.null(n <- options$linewidth)) {
        x = knitr:::split_lines(x)
        # any lines wider than n should be wrapped
        if (any(nchar(x) > n)) x = strwrap(x, width = n)
        x = paste(x, collapse = '\n')
      }
      hook_output(x, options)
    })
```

## Q1
**A**.  

*Compute the OR for actn3_rs540874*
```{r}
    source("https://raw.githubusercontent.com/xuranw/469_public/master/hw2/hw2_functions.R")
    famuss <- read.csv("https://raw.githubusercontent.com/xuranw/469_public/master/hw2/synthetic_famuss.csv")
    or_df <- famuss %>% subset(., actn3_rs540874 != 2) %>%
             select(actn3_rs540874, heart_disease)
    D_A <- table(or_df)[2,2]
    H_A <- table(or_df)[2,1]
    D_a <- table(or_df)[1,2]
    H_a <- table(or_df)[1,1]
    (D_A / H_A) /  (D_a / H_a)
```
*Interpretation*
Observe that $OR \approx 1$, which suggests that the risk factor actn3_rs540874 (in isolation) is not significantly associated with the heart disease. 

**B**.  
*Use logistic regression with the glm function to regress only
heart disease onto all 79 SNP’s (i.e., not including the variable norm_BMI)*
```{r}
    glm_famuss <- famuss[, -which(colnames(famuss) == "norm_BMI")] 
    glm_res <- stats::glm(heart_disease ~ . , data = glm_famuss, family = stats::binomial)
```

_How many SNPs are present in the dataset?_  
```{r}
    length(setdiff(colnames(famuss),  c("id", "Gender", "Age", "Race")))
```
_Using the which function, determine which column in famuss contains information about the the `actn3 1671064` SNP._ 
```{r}
    which(colnames(famuss) == "actn3_1671064")
```
_Using the table function, determine its minor genotype and its frequency_

```{r}
    table(famuss["actn3_1671064"])[1]
```
_Furthermore, using the table function, report these frequencies stratified by the variable Race._  
```{r}
    library(tidyverse)
    famuss %>% select("actn3_1671064", "Race") %>% table
```

**C**.  
```{r}
    allele_table  <- table(famuss["esr1_rs2077647"])
    names(allele_table) <- c(1, 0, 2)
    allele_df <- data.frame(allele_table)
    ggplot(allele_df, aes(x = Var1, y = Freq)) + 
        geom_bar(stat = "identity")  + 
        labs(title = "Genotype Counts of SNP esr1_rs2077647") +
        xlab( "Numeric coding for SNP esr1_rs2077647") +
        ylab("Count")
```  

## Q2
**A**.  
```{r}
    source("https://raw.githubusercontent.com/xuranw/469_public/master/hw1/clt.R")
```
`generate_data` samples $n$ elements from the set ${1,2,3}$ (with replacement). The number of $1$'s in the sample 
determines the number of Gaussian ($\mu = 10, \sigma = 1$), random variables which will be in `sample_vec`.
Similarly, the number of $2$'s in the sample determines the number of Gamma ($\kappa = 2, \theta = 2$) 
random variables which will be in `sample_vec`. Same story with Chi Squared. Our result is a vector of vector of size 
$n$ with three random variables interleaved. 


**B**.  

```{r}
    M <- matrix(rep(NA, 6 * 10000), nrow = 10000)
    n_args <- c(1,2,3,5,10,100)
    i <- 1
    sample_vec <- c(NA, 10000)
    for (val in n_args) {
        M[, i] <- replicate(10000, mean(generate_data(val)))
        i <- i + 1
    }
```
```{r}
    par(mfrow = c(2,3))
    n_1_df <-  data.frame(M[,1])
    hist(n_1_df$M...1., breaks = 100, border= "red",
         col= "green", main = "Sample size: 1",
         xlab = "Value", ylab =  "Frequency")
    n_2_df <-  data.frame(M[,2])
    hist(n_2_df$M...2., breaks = 100, border= "red",
         col= "green", main = "Sample size: 2",
         xlab = "Value", ylab =  "Frequency")
    n_3_df <-  data.frame(M[,3])
    hist(n_3_df$M...3., breaks = 100, border= "red",
         col= "green", main = "Sample size: 3",
         xlab = "Value", ylab =  "Frequency")
    n_5_df <-  data.frame(M[,4])
    hist(n_5_df$M...4., breaks = 100, border= "red",
         col= "green", main = "Sample size: 5",
         xlab = "Value", ylab =  "Frequency")
    n_10_df <-  data.frame(M[,5])
    hist(n_10_df$M...5., breaks = 100, border= "red",
         col= "green", main = "Sample size: 10",
         xlab = "Value", ylab =  "Frequency")
    n_100_df <-  data.frame(M[,6])
    hist(n_100_df$M...6., breaks = 100, border= "red",
         col= "green", main = "Sample size: 100",
         xlab = "Value", ylab =  "Frequency")
```

We note that for $n \leq 3$ (and $n = 5$ to a certain extent) the distribution of means of `sample_vec` 
is multimodal since our sample is too small to converge to anything. For $n \ge 10$ we see our histogram approximate 
the famous Bell curve since our statistic (The mean in this case) is additive with respect to the Random Variables in 
question, which verifies CLT (Lyapunov version specifically since our Random Variables are not Identically distributed).
